Estimating duration using survival analysis

### 2024-07-31 cis-perfect-testing.tex
Rob second p87 comment

@@

Now define the probabilities of each possible outcome of $O(j)$.
For detected outcomes, \ie $\vec{\nu}_k \in \set{E}$, let $p_k = \prob(O(j) = \vec{\nu}_k \mid \vec{\theta})$.
For undetected outcomes, let $p_u = \prob(O(j) = \emptyset \mid \vec{\theta})$.
Together, $\vec{p} = [p_1, \dots, p_{N_E}, p_u]^T$.
Assume, for tractability, that the events $O(j) = \vec{\nu}_k$ and $O(j') = \vec{\nu}_{k'}$ are independent for $j \neq j'$; this assumption is discussed in \cref{perf-test:sec:discussion}.

Then:
\begin{align}
  \vec{n} \mid \ntot, \vec{\theta} &\dist \MN(\ntot, \vec{p})
\intertext{that is:}
  p(\vec{n} \mid \ntot, \vec{\theta}) &= \frac{\ntot!}{\nnodet!\prod_{k=1}^{N_E} n_k!} p_u^{\nnodet} \prod_{k=1}^{N_E} p_k^{n_k}.
  \label{perf-test:eq:multinomial-ll}
\end{align}
\begin{figure}
\makebox[\textwidth][c]{\includegraphics[width=0.9\paperwidth]{cis-perfect-testing/regions_diag}}
\thisfloatpagestyle{empty}
\caption[Episode regions]{%
  Each dot is a combination of $b_j$ and $e_j$ for an arbitrary individual $i$.
  Each box, bounded by dashed lines, are combinations giving $O(j) = \vec{\nu}_k$, $k$ corresponding to the numeric label; labels are arbitrary but unique across all individuals.
  $i$ had negative tests at times 0, 7, 14, 56, and 84 (not shown) and positive tests at times 21 and 28.
  The purple region corresponds to the doubly interval censored episode in this individual.
  That is, $n_8 = 1$ and $n_k = 0$ for $k = 1, \dots, 7, 9, \dots, 15$.
  The red region corresponds to combinations giving $O(j) = \emptyset$.
  Impossible region violates $b_j \leq e_j$.
}
\label{perf-test:fig:partitionSpace}
\end{figure}

@@

Decompose $p_k$ as $p_k = p_{ik} \prob(i(j) = i_k \mid \vec{\theta})$
where $p_{ik} = \prob(O(j) = \vec{\nu}_k \mid i(j) = i_k, \vec{\theta})$.
% This is valid as $\prob(O(j) = \nu_k \mid i(j) \neq i_k) = 0$ due to the condition here being equivalent to equating the two vectors' final elements.
Assume that each infection episode occurs independently and with equal probability in any individual, \ie $\prob(i(j) = i_k) = 1/\Ncis$ for all $j$ and $k$.

If $i(j) = i_k$ then the event $O(j) = \vec{\nu}_k$ occurs if and only if the episode starts in the interval $[l^{(b)}_k, r^{(b)}_k]$ and ends in the interval $[l^{(e)}_k, r^{(e)}_k]$.
Omitting the conditioning on $\vec{\theta}$ and $i(j) = i$, this gives:
\begin{align}
p_{ik}
=& \prob \left( l_k^{(b)} \leq B_{j} \leq r_k^{(b)}, l_k^{(e)} \leq E_{j} \leq r_k^{(e)} \right) \\
=& \prob \left( l_k^{(e)} \leq E_{j} \leq r_k^{(e)} \mid l_k^{(b)} \leq B_{j} \leq r_k^{(b)} \right) \times\prob \left( l_k^{(b)} \leq B_{j} \leq r_k^{(b)} \right) \\
=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \prob \left( l_k^{(e)} \leq E_{j} \leq r_k^{(e)} \mid B_{j} = b \right) \prob \left(B_{j} = b \right) \\
=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \prob \left( l_k^{(e)} - b + 1 \leq D_{j} \leq r_k^{(e)} - b + 1 \right) \prob \left(B_{j} = b \right) &\text{by def of $D_{j}$} \\
=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right) \prob \left(B_{j} = b \right) &\text{by def of $S_{\vec{\theta}}$} \\
\propto& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right)
\label{perf-test:eq:pia}
\end{align}
under the assumption of uniform probability of infection time.
This is the standard form of the likelihood for doubly interval censored data without truncation~\autocite[e.g.][]{sunEmpirical}.

@@

The derivation in \cref{perf-test:sec:model} relied on an assumption that $O(j)$ and $O(j')$ occur independently for $j \neq j'$.
This is not the case, as immunity means that an individual previously infected is less likely to be infected in the future.
Additionally, the same individual cannot have multiple detected concurrent infections, due to the way that the episodes dataset is constructed (see \cref{E-biology-data:sec:cis-episodes}); this limits certain combinations from occurring.
In practice, this means that $n_k > 1$ cannot occur.
The simulation study violated this assumption to a large extent: each individual could only have one infection.
Despite this, the method performed well, recovering the true survival function.
Therefore, it is unlikely that this assumption matters, possibly due to the large number of individuals without detected episodes.
Further work could explore different assumptions, \eg a ``full immunity'' assumption meaning each individual has zero or one episode in the period.

>>>>

For our purposes, an infection episode $j$ consists of the three random variables: $i(j)$, the individual it occurs in; $B_j$, the day it begins (the first day $i(j)$ is detectable due to this infection episode); $E_j$, the day it ends (the last day $i(j)$ is detectable due to this infection episode); the distribution of the episode's duration is a deterministic transformation of these variables, $D_j = E_j - B_j + 1$.
I assume that $B_j$, $D_j$, and $i(j)$ are independent.
$D_j$'s distribution is determined by $\vec{\theta}$; both $B_j$ and $i(j)$ have uniform priors over their respective state spaces.
$O(j)$ is a deterministic function of $B_j$, $E_j$, and $\sched_{i(j)}$.

To make this problem tractable, I model infection episodes as independent conditional on $\ntot$, the total number of them that occur, and $\vec{\theta}$; \ie $(B_j, E_j, i(j)) \ind (B_{j'}, E_{j'}, i(j')) \mid \ntot, \vec{\theta}$ for $j \neq j'$.
This is true both for infection episodes within the same individual and across individuals; this assumption is discussed in \cref{perf-test:sec:discussion}.

@@

Now define the probabilities of each possible outcome of $O(j)$.
For detected outcomes, \ie $\vec{\nu}_k \in \set{E}$, let $p_k = \prob(O(j) = \vec{\nu}_k \mid \vec{\theta})$.
For undetected outcomes, let $p_u = \prob(O(j) = \emptyset \mid \vec{\theta})$.
Together, $\vec{p} = [p_1, \dots, p_{N_E}, p_u]^T$.

As we assume $O(j) = \vec{\nu}_k$ and $O(j') = \vec{\nu}_{k'}$ are independent for $j \neq j'$:
\begin{align}
  \vec{n} \mid \ntot, \vec{\theta} &\dist \MN(\ntot, \vec{p})
\intertext{that is:}
  p(\vec{n} \mid \ntot, \vec{\theta}) &= \frac{\ntot!}{\nnodet!\prod_{k=1}^{N_E} n_k!} p_u^{\nnodet} \prod_{k=1}^{N_E} p_k^{n_k}.
  \label{perf-test:eq:multinomial-ll}
\end{align}
\begin{figure}
\makebox[\textwidth][c]{\includegraphics[width=0.9\paperwidth]{cis-perfect-testing/regions_diag}}
\thisfloatpagestyle{empty}
\caption[Episode regions]{%
  Each dot is a combination of $b_j$ and $e_j$ for an arbitrary individual $i$.
  Each box, bounded by dashed lines, are combinations giving $O(j) = \vec{\nu}_k$, $k$ corresponding to the numeric label; labels are arbitrary but unique across all individuals.
  $i$ had negative tests at times 0, 7, 14, 56, and 84 (not shown) and positive tests at times 21 and 28.
  The purple region corresponds to the doubly interval censored episode in this individual.
  That is, $n_8 = 1$ and $n_k = 0$ for $k = 1, \dots, 7, 9, \dots, 15$.
  The red region corresponds to combinations giving $O(j) = \emptyset$.
  Impossible region violates $b_j \leq e_j$.
}
\label{perf-test:fig:partitionSpace}
\end{figure}

@@

Decompose $p_k$ as $p_k = p_{ik} \prob(i(j) = i_k \mid \vec{\theta})$
where $p_{ik} = \prob(O(j) = \vec{\nu}_k \mid i(j) = i_k, \vec{\theta})$.
% This is valid as $\prob(O(j) = \nu_k \mid i(j) \neq i_k) = 0$ due to the condition here being equivalent to equating the two vectors' final elements.
As I assume that each infection episode occurs independently and with equal probability in any individual, $\prob(i(j) = i_k) = 1/\Ncis$ for all $j$ and $k$.

If $i(j) = i_k$ then the event $O(j) = \vec{\nu}_k$ occurs if and only if the episode starts in the interval $[l^{(b)}_k, r^{(b)}_k]$ and ends in the interval $[l^{(e)}_k, r^{(e)}_k]$.
Omitting the conditioning on $\vec{\theta}$ and $i(j) = i$, this gives:
\begin{align}
p_{ik}
=& \prob \left( l_k^{(b)} \leq B_{j} \leq r_k^{(b)}, l_k^{(e)} \leq E_{j} \leq r_k^{(e)} \right) \\
=& \prob \left( l_k^{(e)} \leq E_{j} \leq r_k^{(e)} \mid l_k^{(b)} \leq B_{j} \leq r_k^{(b)} \right) \times\prob \left( l_k^{(b)} \leq B_{j} \leq r_k^{(b)} \right) \\
=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \prob \left( l_k^{(e)} \leq E_{j} \leq r_k^{(e)} \mid B_{j} = b \right) \prob \left(B_{j} = b \right) \\
=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \prob \left( l_k^{(e)} - b + 1 \leq D_{j} \leq r_k^{(e)} - b + 1 \right) \prob \left(B_{j} = b \right) &\text{by def of $D_{j}$} \\
=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right) \prob \left(B_{j} = b \right) &\text{by def of $S_{\vec{\theta}}$} \\
\propto& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right)
\label{perf-test:eq:pia}
\end{align}
by the assumption of uniform probability of infection time.
This is the standard form of the likelihood for doubly interval censored data without truncation~\autocite[e.g.][]{sunEmpirical}.

@@

The derivation in \cref{perf-test:sec:model} relied on an assumption that infection episodes occur independently.
This is not the case, as immunity means that an individual previously infected is less likely to be infected in the future.
Additionally, the same individual cannot have multiple detected concurrent infections, due to the way that the episodes dataset is constructed (see \cref{E-biology-data:sec:cis-episodes}); this limits certain combinations from occurring.
In practice, this means that $n_k > 1$ cannot occur.
The simulation study violated this assumption to a large extent: each individual could only have one infection.
Despite this, the method performed well, recovering the true survival function.
Therefore, it is unlikely that this assumption matters, possibly due to the large number of individuals without detected episodes.
Further work could explore different assumptions, \eg a ``full immunity'' assumption meaning each individual has zero or one episode in the period.
